---
title: "Homework 2 Report"
output: html_document
---


## Carly Gaudin

### Setup

Read in the data with `read_csv()` and store the data as an R object named `dataset`. Check the data to make sure all of the expected observations and variables are there.


```{r}
## Load the data and any necessary packages here.
library(readr)
library(ggplot2)
library(dplyr)
dataset <- read_csv('maacs.csv.gz')
```

## Part 1

We will first consider the relationship between FEV1 and age. In general, it is expected that as children get older (and hence, larger in size), their FEV1 values should get higher. 

Consider the statement "FEV1 values in children are higher in older children relative to younger children". 

Write a function in R that takes the `dataset` object as an argument and returns `TRUE` if the statement above is true for the dataset and `FALSE` otherwise. 

NOTE: In order to write this function, you will need to translate the statement above into something that can be checked with the data. There are many ways in which you can do that translation correctly and you only need to pick one way here. 

NOTE: For this part, do not use any plots.

```{r}
## Write your function here
fev1_age_correlation <- function(dataset) {
  # remove rows with NA values for fev1 and age variables
  clean_data <- dataset %>%
    filter(!is.na(fev1) & !is.na(age))
  
  # determine mean age for the whole dataset
  mean_age <- mean(clean_data$age)
  
  # split dataset into older and younger datasets based on mean age
  younger_children <- dataset[clean_data$age <= mean_age, ]
  older_children <- dataset[clean_data$age > mean_age, ]
  
  # determine mean of fev1 for the younger and older dataset
  young_avg_fev1 <- mean(younger_children$fev1, na.rm = TRUE)
  old_avg_fev1 <- mean(older_children$fev1, na.rm = TRUE)
  
  # return TRUE if older children have a higher average fev1 value than     younger children, return FALSE otherwise
  return(old_avg_fev1 > young_avg_fev1)
}
```


## Part 2

Fit a linear regression model with FEV1 as the outcome and age as a predictor.

How much does FEV1 change for a 1-year increase in the child's age? 

```{r}
# fit linear regression model
model <- lm(fev1 ~ age, data = dataset)

# extract slope
coef(model)["age"]
```

Th estimated slope coefficient for age was about 0.17. This means that, on average, for a one-year increase in a child's age, the FEV1 value increases about 0.17 liters.


## Part 3

Develop **three** supporting premises derived from the data that support the statement you wrote in Part 2. These can be plots, other summary statistics, or model results.

NOTE: At least one supporting premise should use a plot.

```{r}
# Supporting premise #1 
# scatterplot using Age and FEV! to show their relationship
ggplot(dataset, aes(x = age, y = fev1)) +
  geom_point(color = "blue") +
  labs(title = "Relationship between Age and FEV1",
       x = "Age (years)", y = "FEV1 (liters)")

# Supporting Premise #2
# determine mean age for the whole dataset
mean_age <- mean(dataset$age, na.rm = TRUE)
  
# mean of fev1 value of the older children (older than the mean of all children)
older_children <- dataset[dataset$age > mean_age, ]
old_avg_fev1 <- mean(older_children$fev1, na.rm = TRUE)
old_avg_fev1

# mean of fev1 value of the younger children (younger or same age of all children)
younger_children <- dataset[dataset$age <= mean_age, ]
young_avg_fev1 <- mean(younger_children$fev1, na.rm = TRUE)
young_avg_fev1

# Supporting Premise #3
# correlation value of age and fev1 values, disregarding NA values
cor(dataset$age, dataset$fev1, use = "complete.obs")
```


Write the three supporting premise statements here:

1. My scatterplot of Age vs. FEV1 shows a positive trend, where older children typically have a higher FEV1 value. This evidence supports a positive realtionship betwee age and FEV1.

2. When splitting the children into two groupd based on the mean age, the average FEV1 value of older children is higher than the average FEV1 value of younger children. This group comparison reinforces the idea that older children have higher FEV1 values.

3. The correlation coefficient between age and FEV1 is postive, confirming that as age increase, FEV1 typically increases as well. 


## Part 4

For each of the supporting premises above, write a function that takes the `dataset` object as an argument and returns `TRUE` if the supporting premise statement above is true for the dataset and `FALSE` otherwise. 

For statements involving plots, instead of returning `TRUE` or `FALSE`, you function should do two things:

1. Produce the plot that is used in the statement

2. Produce a hypothetical version of the plot in the event that the statement is true. This can be done using simulated data or by simply hand drawing a plot.


```{r}
## Function for supporting premise statement 1
# plot showing relationship between age and fev1
supporting_premise1 <- function(dataset) {
  library(ggplot2)
  
  # Actual plot
  p1 <- ggplot(dataset, aes(x = age, y = fev1)) +
    geom_point(color = "blue") +
    labs(title = "Observed Relationship between Age and FEV1",
         x = "Age (years)", y = "FEV1 (liters)")
  print(p1)
  
  # hypothetical plot (simulated upward trend)
  sim_data <- data.frame(
  age = seq(min(dataset$age, na.rm = TRUE), 
            max(dataset$age, na.rm = TRUE), length.out = 100),
  fev1 = seq(min(dataset$fev1, na.rm = TRUE), 
             max(dataset$fev1, na.rm = TRUE), length.out = 100) + rnorm(100, 0, 0.1))
  
  p2 <- ggplot(sim_data, aes(x = age, y = fev1)) +
    geom_point(color = "blue") +
    labs(title = "Hypothetical Positive Relationship (if premise is true)",
         x = "Age (years)", y = "FEV1 (liters)")
  print(p2)}
```


```{r}
## Function for supporting premise statement 2
# compare mean fev1 of older vs younger children
supporting_premise2 <- function(dataset) {
  mean_age <- mean(dataset$age, na.rm = TRUE)
  
  old_avg_fev1 <- mean(dataset$fev1[dataset$age > mean_age], na.rm = TRUE)
  young_avg_fev1 <- mean(dataset$fev1[dataset$age <= mean_age], na.rm = TRUE)
  
  return(old_avg_fev1 > young_avg_fev1) }
```


```{r}
## Function for supporting premise statement 3
# check correlation between age and fev1
supporting_premise3 <- function(dataset) {
  correlation <- cor(dataset$age, dataset$fev1, use = "complete.obs")
  
  return(correlation > 0)}
```

Execute each of your function and show that the produce the expected output.
```{r}
# running each premise function
supporting_premise1(dataset)   # produces observed + hypothetical plots
supporting_premise2(dataset)   # returns TRUE or FALSE
supporting_premise3(dataset)   # returns TRUE or FALSE

```



## Part 5

Describe one alternative to the primary statement "FEV1 values in children are higher in older children relative to younger children". 

Create a fault tree for the alternative outcome describing how the alternative outcome could be realized in the data even if the primary statement were true.

Your fault tree should be created as a separate image and does not need to be created in R. Upload the image of the fault tree to Canvas.


